Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans

### 2020-06-03 paper_PLOS/paper_PLOS_draft.tex
Fixed Figures and converted to bookdown-based knitting, need to fix HTML and WORD outputs

% Pandoc header
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Our initial hypothesis was that genes which duplicated in lineages that experienced a growth in size would be enriched for membership in tumor suppressor pathways. Thus, we used WebGestalt and its Overrepresentation Analysis (ORA) functionality to determine what pathways were enriched in our duplicated gene sets in each branch relative to our initial query set. For our database, we used Reactome for our primary analysis, but additionally used the KEGG, Panther, Wikipathways, and Wikipathways\_cancer databases using WebGestalt ORA. Going through the tree, at no FDR \(\leq\) 0.5 is there any significant pathway representation for genes that increased in the branches leading to \emph{Afrosoricida}, \emph{Pseudoungulata}, \emph{Elephantina}, \emph{Loxodontini}, \emph{Elephas maximus}, \emph{Mammuthus}, \emph{Loxodonta cyclotis}, \emph{Palaeoloxodon antiquus}, \emph{Loxodona}, \emph{Mammuthus columbi}, \emph{Mammuthus primigenius}, \emph{Procavia capensis}, \emph{Elephantidae}, or \emph{Proboscidea}; furthermore, there are no significant pathway enrichments at FDR\textless{}0.5 for genes whose copy number did not change between branches (copy-number-stable) for \_Afrotheria\_\_. Note that because \emph{Xenarthra} was selected as the outgroup, it is not possible to polarize the changes in their gene copy numbers along the tree.

@@

\begin{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-intro-1} \hfill{}

\caption{Atlantogenatans with sequenced genomes, body sizes, and known lifespans [@PuttickAndThomas2015; @HAGR]}\label{fig:Figure-intro}
\end{figure}

\begin{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-BodySize-1} \hfill{}

\caption{Body sizes rapidly and frequently expand in Eutherians, especially in Atlantogenata. **A)** Tree of Eutherian species, colored by ln(Body Size) and with branch lengths set to the rate of change in body sizes, normalized by the square root of the root branch. Atlantogenata is highlighted at the bottom. **B)** Zoom-in of (**A**) on Atlantogenata. Silhuetes for the African Elephant, West Indian Manatee, Cape Elephant Shrew, Lesser Hedgehog Tenrec, Cape Golden Mole, Nine-Banded Armadillo, and Hoffman's Two-Toed Sloth are colored by their extant body sizes, while clade labels are colored based on the common ancestor's estimated body size. **C)** Confidence interval plot for representative species and ancestral nodes.}\label{fig:Figure-BodySize}
\end{figure}

\begin{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-CancerSuscept-1} \hfill{}

\caption{Cancer succecptibility across Atlantogenata. Branch lengths are set to the magnitude of change in cancer succeptiblity; colors indicate the magnitude and direction of the change.}\label{fig:Figure-CancerSuscept}
\end{figure}

\begin{figure}

\includegraphics[width=6in,height=7in]{paper_PLOS_draft_files/figure-latex/Figure-GeneDup-Cladogram-1} \hfill{}

\caption{Gene duplications occur readily throughout Atlantogenata. Shown here is a tree of Atlantogenatan species with genomes, with the number of genes that underwent an increase in copy number overlayed at each node.}\label{fig:Figure-GeneDup-Cladogram}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=6in,height=6in]{paper_PLOS_draft_files/figure-latex/Figure-GenomeCor-1}

@@

\begin{figure}

\includegraphics[width=6in,height=4in]{paper_PLOS_draft_files/figure-latex/Figure-loxAfrORA-1} \hfill{}

\caption{Overrepresentation Analysis of Duplicated Genes in Atlantogenata using Reactome Pathways.}\label{fig:Figure-loxAfrORA}
\end{figure}

\begin{figure}

\includegraphics[width=6in,height=6in]{paper_PLOS_draft_files/figure-latex/Figure-loxAfrTP53-1} \hfill{}

\end{figure}

@@

\caption{\label{tab:Table-SRAData}NCBI SRA datasets used in this study, along with key biological and genome information.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lll>{\raggedright\arraybackslash}p{50em}>{\raggedright\arraybackslash}p{50em}}
\toprule
Organism & Common Name & Genome & SRA Acc. & Tissues\\
\midrule
Dasypus novemcinctus & Nine-banded armadillo & dasNov3 & SRR494779, SRR494767, SRR494780, SRR494770, SRR309130, SRR494771, SRR4043756, SRR494776, SRR494778, SRR4043762, SRR4043755, SRR6206923, SRR4043761, SRR4043760, SRR6206913, SRR4043763, SRR494772, SRR494781, SRR494774, SRR494777, SRR494775, SRR4043754, SRR1289524, SRR4043758, SRR6206903, SRR1289523, SRR4043759, SRR3222425, SRR494768, SRR494769, SRR6206908, SRR4043757, SRR494766, SRR6206918, SRR494773 & Kidney, Spleen, Cerebellum W/ Brainstem, Rt. Quadricep, Mid-Stage Pregnant Endometrium, Cervix, Lung, Liver, Skeletal Muscle, Ascending Colon, Pregnant Armadillo Endometrium, Heart, Placenta\\
Loxodonta africana & African savanna elephant & loxAfr3, loxAfrC, loxAfr4 & SRR6307198, SRR1041765, SRR6307199, SRR6307201, SRR6307196, SRR6307202, SRR6307200, SRR6307195, SRR975188, SRR6307194, SRR6307204, SRR3222430, SRR6307205, SRR975189, SRR6307197, SRR6307203 & Blood, Fibroblast, Placenta\\
Trichechus manatus latirostris & Manatee & triMan1, triManLat2 & SRR4228542, SRR4228545, SRR4228544, SRR4228539, SRR4228541, SRR4228538, SRR4228546, SRR4228537, SRR4228540, SRR4228543, SRR4228547 & Buffy Coat\\
\bottomrule
\end{tabular}}
\end{table}

@@

@@

\textbackslash{}begin\{figure\}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-ECNC-1} \hfill{}

\textbackslash{}caption\{Estimated Copy Number by Coverage (ECNC) consolidates fragmented genes while accounting for missing domains in homologs. \textbf{A)} A single, contiguous gene homolog in a target genome with 100\% query length coverage has an ECNC of 1.0. \textbf{B)} Two contiguous gene homologs, each with 100\% query length coverage have an ECNC of 2.0. \textbf{C)} A single gene homolog, split across multiple scaffolds and contigs in a fragmented target genome; BLAT identifies each fragment as a single hit. Per nucleotide of query sequence, there is only one corresponding nucleotide over all the hits, thus the ECNC is 1.0. \textbf{D)} Two gene homologs, one fragmented and one contiguous. 100\% of nucleotides in the query sequence are represented between all hits; however, every nucleotide in the query has two matching nucleotides in the target genome, thus the ECNC is 2.0. \textbf{E)} One true gene homolog in the target genome, plus multiple hits of a conserved domain that span 20\% of the query sequence. While 100\% of the query sequence is represented in total, 20\% of the nucleotides have 4 hits. Thus, the ECNC for this gene is 1.45. \textbf{F)} Two real gene homologs; one hit is contiguous, one hit is fragmented in two, and the tail end of both sequences was not identified by BLAT due to sequence divergence. Only 75\% of the query sequence was covered in total between the hits, but for that 75\%, each nucleotide has two hits. As such, ECNC is equal to 2.0 for this gene.\}\label{fig:Figure-ECNC}
\textbackslash{}end\{figure\}

\begin{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-GeneDup-BLScaled-1} \hfill{}

\caption{Gene copy increases polarized along Atlantogenata, colored by ln(Body Size), with branch lengths equal to the change in gene copy number}\label{fig:Figure-GeneDup-BLScaled}
\end{figure}

@@

\begin{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-RBHBStrategy-1} \hfill{}

\caption{Full version of RecBlat strategy, but low quality}\label{fig:Figure-RBHBStrategy}
\end{figure}

\begin{figure}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure-TimeTree-Eutheria-1} \hfill{}

\caption{The time-calibrated Eutherian tree used as input for StableTraits, from Bininda-Emonds et al and Puttick and Thomas}\label{fig:Figure-TimeTree-Eutheria}
\end{figure}

>>>>

% Pandoc header
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}

\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Our initial hypothesis was that genes which duplicated in lineages that experienced a growth in size would be enriched for membership in tumor suppressor pathways. Thus, we used WebGestalt and its Overrepresentation Analysis (ORA) functionality to determine what pathways were enriched in our duplicated gene sets in each branch relative to our initial query set. For our database, we used Reactome for our primary analysis, but additionally used the KEGG, Panther, Wikipathways, and Wikipathways\_cancer databases using WebGestalt ORA. Going through the tree, at no FDR \(\leq\) 0.5 is there any significant pathway representation for genes that increased in the branches leading to \emph{Afrosoricida}, \emph{Pseudoungulata}, \emph{Elephantina}, \emph{Loxodontini}, \emph{Elephas maximus}, \emph{Mammuthus}, \emph{Loxodonta cyclotis}, \emph{Palaeoloxodon antiquus}, \emph{Loxodona}, \emph{Mammuthus columbi}, \emph{Mammuthus primigenius}, \emph{Procavia capensis}, \emph{Elephantidae}, or \emph{Proboscidea}; furthermore, there are no significant pathway enrichments at FDR\textless{}0.5 for genes whose copy number did not change between branches (copy-number-stable) for \emph{Afrotheria}. Note that because \emph{Xenarthra} was selected as the outgroup, it is not possible to polarize the changes in their gene copy numbers along the tree.

@@

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-intro-1} \caption{\emph{Atlantogenatans} with sequenced genomes, body sizes, and known lifespans {[}14,21{]}}\label{fig:Figure-intro}
\end{figure}

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-BodySize-1} \caption{Body sizes rapidly and frequently expand in Eutherians, especially in Atlantogenata. **A)** Tree of Eutherian species, colored by ln(Body Size) and with branch lengths set to the rate of change in body sizes, normalized by the square root of the root branch. Atlantogenata is highlighted at the bottom. **B)** Zoom-in of (**A**) on Atlantogenata. Silhuetes for the African Elephant, West Indian Manatee, Cape Elephant Shrew, Lesser Hedgehog Tenrec, Cape Golden Mole, Nine-Banded Armadillo, and Hoffman's Two-Toed Sloth are colored by their extant body sizes, while clade labels are colored based on the common ancestor's estimated body size. **C)** Confidence interval plot for representative species and ancestral nodes.}\label{fig:Figure-BodySize}
\end{figure}

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-CancerSuscept-1} \caption{Cancer succecptibility across \emph{Atlantogenata.} Branch lengths are set to the magnitude of change in cancer succeptiblity; colors indicate the magnitude and direction of the change.}\label{fig:Figure-CancerSuscept}
\end{figure}

\begin{figure}[H]
\includegraphics[width=6in,height=6in,]{paper_PLOS_draft_files/figure-latex/Figure-GeneDup-Cladogram-1} \caption{Gene duplications occur readily throughout \emph{Atlantogenata}. Shown here is a tree of \emph{Atlantogenatan} species with genomes, with the number of genes that underwent an increase in copy number overlayed at each node.}\label{fig:Figure-GeneDup-Cladogram}
\end{figure}

\begin{figure}[H]

{\centering \includegraphics[width=6in,height=6in,]{paper_PLOS_draft_files/figure-latex/Figure-GenomeCor-1}

@@

\begin{figure}[H]
\includegraphics[width=6in,height=4in,]{paper_PLOS_draft_files/figure-latex/Figure-loxAfrORA-1} \caption{Overrepresentation Analysis of Duplicated Genes in Atlantogenata using Reactome Pathways.}\label{fig:Figure-loxAfrORA}
\end{figure}

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-loxAfrTP53-1} \caption{TP53-related genes are also duplicated and functional in \emph{Loxodonta africana}. \textbf{A)} Cladogram of \emph{Atlantogenata} highlighting along each branch when duplications for each gene occured. \textbf{B)} Gene expression data collected from pulically-available RNA-seq data for each duplicate in \textbf{A)}.}\label{fig:Figure-loxAfrTP53}
\end{figure}

@@

@@

\caption{\label{tab:Table-SRAData}NCBI SRA datasets used in this study, along with key biological and genome information.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lll>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{10em}}
\toprule
Organism & Common Name & Genome & SRA Acc. & Tissues\\
\midrule
Dasypus novemcinctus & Nine-banded armadillo & dasNov3 & SRR494779, SRR494767, SRR494780, SRR494770, SRR309130, SRR494771, SRR4043756, SRR494776, SRR494778, SRR4043762, SRR4043755, SRR6206923, SRR4043761, SRR4043760, SRR6206913, SRR4043763, SRR494772, SRR494781, SRR494774, SRR494777, SRR494775, SRR4043754, SRR1289524, SRR4043758, SRR6206903, SRR1289523, SRR4043759, SRR3222425, SRR494768, SRR494769, SRR6206908, SRR4043757, SRR494766, SRR6206918, SRR494773 & Kidney, Spleen, Cerebellum W/ Brainstem, Rt. Quadricep, Mid-Stage Pregnant Endometrium, Cervix, Lung, Liver, Skeletal Muscle, Ascending Colon, Pregnant Armadillo Endometrium, Heart, Placenta\\
Loxodonta africana & African savanna elephant & loxAfr3, loxAfrC, loxAfr4 & SRR6307198, SRR1041765, SRR6307199, SRR6307201, SRR6307196, SRR6307202, SRR6307200, SRR6307195, SRR975188, SRR6307194, SRR6307204, SRR3222430, SRR6307205, SRR975189, SRR6307197, SRR6307203 & Blood, Fibroblast, Placenta\\
Trichechus manatus latirostris & Manatee & triMan1, triManLat2 & SRR4228542, SRR4228545, SRR4228544, SRR4228539, SRR4228541, SRR4228538, SRR4228546, SRR4228537, SRR4228540, SRR4228543, SRR4228547 & Buffy Coat\\
\bottomrule
\end{tabular}}
\end{table}

@@

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-ECNC-1} \caption{Estimated Copy Number by Coverage (ECNC) consolidates fragmented genes while accounting for missing domains in homologs. \textbf{A)} A single, contiguous gene homolog in a target genome with 100\% query length coverage has an ECNC of 1.0. \textbf{B)} Two contiguous gene homologs, each with 100\% query length coverage have an ECNC of 2.0. \textbf{C)} A single gene homolog, split across multiple scaffolds and contigs in a fragmented target genome; BLAT identifies each fragment as a single hit. Per nucleotide of query sequence, there is only one corresponding nucleotide over all the hits, thus the ECNC is 1.0. \textbf{D)} Two gene homologs, one fragmented and one contiguous. 100\% of nucleotides in the query sequence are represented between all hits; however, every nucleotide in the query has two matching nucleotides in the target genome, thus the ECNC is 2.0. \textbf{E)} One true gene homolog in the target genome, plus multiple hits of a conserved domain that span 20\% of the query sequence. While 100\% of the query sequence is represented in total, 20\% of the nucleotides have 4 hits. Thus, the ECNC for this gene is 1.45. \textbf{F)} Two real gene homologs; one hit is contiguous, one hit is fragmented in two, and the tail end of both sequences was not identified by BLAT due to sequence divergence. Only 75\% of the query sequence was covered in total between the hits, but for that 75\%, each nucleotide has two hits. As such, ECNC is equal to 2.0 for this gene.}\label{fig:Figure-ECNC}
\end{figure}

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-GeneDup-BLScaled-1} \caption{Gene copy increases polarized along Atlantogenata, colored by ln(Body Size), with branch lengths equal to the change in gene copy number}\label{fig:Figure-GeneDup-BLScaled}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-RBHBStrategy-1} \caption{Full version of RecBlat strategy, but low quality}\label{fig:Figure-RBHBStrategy}
\end{figure}

\begin{figure}[H]
\includegraphics[width=6in,]{paper_PLOS_draft_files/figure-latex/Figure-TimeTree-Eutheria-1} \caption{The time-calibrated Eutherian tree used as input for StableTraits, from Bininda-Emonds et al and Puttick and Thomas}\label{fig:Figure-TimeTree-Eutheria}
\end{figure}
